A Generalization of Regularized Discriminant Analysis for High-Dimensional Classification

### 2013-03-24 reports/paper-rda.tex
Added overview of simulation design.

Microarrays have enabled researchers to measure gene-expression levels of thousands of genes in a single experiment to identify a subset of genes and their expression patterns that either characterize a particular cell state or facilitate an accurate prediction of future cell state.  We focus on the classification of tissues and other biological samples into one of $K$ disjoint, predefined cancer types, where our goal is to ascertain the class membership of a new tissue sample based on gene-expression microarray data. However, microarray data typically consist of a large number of gene expressions $p$ relative to the sample size $N$. Moreover, standard benchmark microarray data sets usually consist of between $p = 10,000$ and $55,000$ probes and fewer than $N = 100$ observations, in which case standard supervised-learning methods exhibit poor classification performance, due to the \emph{curse of dimensionality} \citep{Bellman:1961tn}. Furthermore, well-known classifiers, including linear discriminant analysis \emph{(LDA)} and quadratic discriminant analysis \emph{(QDA)}, are incalculable when $p > N$ because the sample covariance matrices employed in both classifiers are singular. As a result, many researchers, including \cite*{Merchante:2012vk}, \cite{Witten:2011kc}, \cite*{Pang:2009ik}, \cite*{Huang:2010ju}, and \cite*{Clemmensen:2011kr}, have introduced restrictions on the population covariance matrices in the \emph{LDA} and \emph{QDA} models to reduce number of covariance parameters that must be estimated to ensure that the resulting classifiers are calculable.  For instance, \cite{Dudoit:2002ev} have proposed the diagonal \emph{LDA} (\emph{DLDA}) classifier, a modification of the \emph{LDA} classifier where the class covariance matrices are assumed to be diagonal.

@@

Thus far, we have construction each $\tilde{\bm \Sigma}_k$ in terms of the first $q$ eigenvectors of $\widehat{\bm \Sigma}$. In practice, we would usually require a costly eigendecomposition of $\widehat{\bm \Sigma}$, but by utilizing a technique similar to the so-called kernel trick, we can obtain $\bm U_1$ by computing the eigendecomposition of a much smaller $N \times N$ matrix when $p \gg N$ \citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix $\widehat{\bm\Sigma} = \bm X' \bm X$. Then, by writing the SVD of $\bm X = \bm M \bm D^{1/2} \bm U'$, we have the spectral decomposition $\widehat{\bm\Sigma} = \bm U \bm D \bm U'$, where $\bm M \in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N \times N}$ is a diagonal matrix consisting of the singular values of $\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, we see that $\bm X \bm X' = \bm M \bm D \bm M'$ is the eigenvalue decomposition of $\bm X \bm X'$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X'$ and the columns of $\bm M$  are the corresponding eigenvectors. Hence, we obtain $\bm M$ and $\bm D$ from an eigendecomposition of $\bm X \bm X' \in \mathbb{R}_{N \times N}$. Next, we compute $\bm U = \bm X' \bm M \bm D^{+/2}$, where $\bm D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We determine $q$ efficiently by computing the number of numerically nonzero eigenvalues present in $\bm D$ such that $q$ is the number of eigenvalues that exceeds some tolerance value, say, $1 \times 10^{-6}$. After calculating $q$, we retain only the first $q$ columns of $\bm U$ to extract $\bm U_1$.

\section{Monte Carlo Simulations}

Our generalization of the \emph{RDA} classifier facilitates a large number of potential models based on the specification of $\alpha_k$ and $\bm \Phi_k$ in \eqref{eq:grda-cov} to shrink the eigenvalues of $\widehat{\bm \Sigma}_k(\lambda)$. In our simulations below, we consider two choices of $\alpha_k$ and $\bm \Phi_k$ based on variants of eigenvalue adjustment methods proposed in the literature. First, ...

Second, \cite{Srivastava:2007ww} derive an empirical-Bayes covariance-matrix estimator under the assumption that the data are independent and identically distributed multivariate normal observations and then substitute the estimator into the \emph{LDA} classifier to attain the minimum distance empirical Bayes (\emph{MDEB}) classifier. To derive the empirical-Bayes covariance-matrix estimator, \cite{Srivastava:2007ww} first assume that $\bm \Sigma^{-1}$ follows a Wishart distribution \emph{a priori} with mean $\lambda^{-1} \bm I_p$, $\lambda > 0$, and degrees of freedom, $l \ge p$. The resulting posterior estimator for $\bm \Sigma^{-1}$ is $\frac{N + l}{N}(N^{-1}\lambda \bm I_p + \widehat{\bm \Sigma})^{-1}$. Using an empirical Bayes argument to estimate $\lambda$, \cite{Srivastava:2007ww} propose the covariance-matrix estimator
\begin{align*}
	\widehat{\bm \Sigma}_{MDEB} = \widehat{\bm \Sigma} + \frac{\tr \{\widehat{\bm \Sigma}\}}{\min\{N, p\}} \bm I_p,
\end{align*}
where the shrinkage factor is approximately equal to the average of the nonzero eigenvalues of $\widehat{\bm \Sigma}$. We consider a variant of the \emph{MDEB} covariance-matrix estimator because \cite{Srivastava:2007ww} claim that the \emph{MDEB} classifier is the best covariance-matrix regularization method applied to the \emph{LDA} classifier they have encountered in the literature.

@@

In this section, we describe three high-dimensional microarray data sets and compare our proposed \emph{GRDA} classifier with two recently proposed classifiers for small-sample, high-dimensional data: the penalized \emph{LDA} (\emph{PLDA}) classifier from \cite{Witten:2011kc}, the shrinkage-mean-based DLDA (\emph{SmDLDA}) classifier from \cite{Tong:2012hw}, the \emph{RDA} classifier as defined in \cite{Hastie:2008dt}, and the sparse discriminant analysis (\emph{SDA}) classifier from \cite{Clemmensen:2011kr}. We calculated 10-fold cross-validation error rates \citep{Hastie:2008dt} to evaluate the performance of each classifier considered. Because the \emph{PLDA} classifier has a built-in variable selection method, for fair comparison we compared each of the classifiers using the variables selected from each cross-validation fold.

@@

\section{Next Actions}

\begin{itemize}
\item Add simulation results
\item Discussion simulation results
\item Show a timing comparison between our implementation and that of the {\tt klaR} package.
\item Discuss choice of $\lambda$ and eigenvalue shrinkage, if necessary
\item Question to answer: What do we do when $\lambda = 0$, corresponding to \emph{QDA}?
\end{itemize}

>>>>

Microarrays have enabled researchers to measure gene-expression levels of thousands of genes in a single experiment to identify a subset of genes and their expression patterns that either characterize a particular cell state or facilitate an accurate prediction of future cell state.  We focus on the classification of tissues and other biological samples into one of $K$ disjoint, predefined cancer types, where our goal is to ascertain the class membership of a new tissue sample based on gene-expression microarray data. However, microarray data typically consist of a large number of gene expressions $p$ relative to the sample size $N$. Moreover, standard benchmark microarray data sets usually consist of between $p = 10,000$ and $55,000$ probes and fewer than $N = 100$ observations, in which case standard supervised-learning methods exhibit poor classification performance, due to the \emph{curse of dimensionality} \citep{Bellman:1961tn}. Furthermore, well-known classifiers, including linear discriminant analysis \emph{(LDA)} and quadratic discriminant analysis \emph{(QDA)}, are incalculable when $p > N$ because the sample covariance matrices employed in both classifiers are singular. As a result, many researchers, including \cite*{Mai:2012bf}, \cite*{Merchante:2012vk}, \cite{Witten:2011kc}, \cite*{Pang:2009ik}, \cite*{Huang:2010ju}, and \cite*{Clemmensen:2011kr}, have introduced restrictions on the population covariance matrices in the \emph{LDA} and \emph{QDA} models to reduce number of covariance parameters that must be estimated to ensure that the resulting classifiers are calculable.  For instance, \cite{Dudoit:2002ev} have proposed the diagonal \emph{LDA} (\emph{DLDA}) classifier, a modification of the \emph{LDA} classifier where the class covariance matrices are assumed to be diagonal.

@@

Thus far, we have constructed each $\tilde{\bm \Sigma}_k$ in terms of the first $q$ eigenvectors of $\widehat{\bm \Sigma}$. By utilizing a technique similar to the so-called kernel trick, we can obtain $\bm U_1$ by computing the eigendecomposition of a much smaller $N \times N$ matrix when $p \gg N$ \citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix $\widehat{\bm\Sigma} = \bm X' \bm X$. Then, by writing the SVD of $\bm X = \bm M \bm D^{1/2} \bm U'$, we have the spectral decomposition $\widehat{\bm\Sigma} = \bm U \bm D \bm U'$, where $\bm M \in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N \times N}^{\ge}$ is a diagonal matrix consisting of the singular values of $\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, we see that $\bm X \bm X' = \bm M \bm D \bm M'$ is the eigenvalue decomposition of $\bm X \bm X'$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X'$ and the columns of $\bm M$  are the corresponding eigenvectors. Hence, we obtain $\bm M$ and $\bm D$ from an eigendecomposition of $\bm X \bm X' \in \mathbb{R}_{N \times N}$. Next, we compute $\bm U = \bm X' \bm M \bm D^{+/2}$, where $\bm D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We determine $q$ efficiently by computing the number of numerically nonzero eigenvalues present in $\bm D$ such that $q$ is the number of eigenvalues that exceeds some tolerance value, say, $1 \times 10^{-6}$. After calculating $q$, we retain only the first $q$ columns of $\bm U$ to extract $\bm U_1$.

@@

In this section, we describe \textbf{three} high-dimensional microarray data sets and compare our proposed \emph{GRDA} classifier with \textbf{three} recently proposed classifiers for small-sample, high-dimensional data: the penalized \emph{LDA} (\emph{PLDA}) classifier from \cite{Witten:2011kc}, the shrinkage-mean-based DLDA (\emph{SmDLDA}) classifier from \cite{Tong:2012hw}, and the sparse discriminant analysis (\emph{SDA}) classifier from \cite{Clemmensen:2011kr}. We evaluated the classification performance of each classifier by randomly partitioning the data such that $2/3$ of the observations were allocated as training data and the remaining $1/3$ of the observations were allocated as a test data set. For each random partition, we applied the feature-selection method from \cite{Dudoit:2002ev} to choose the $d$ features having the largest ratio of their between-group to within-group sums of squares from the training data. We then filtered the test data set with the corresponding $d$ features. After training each classifier from the filtered training data, we classified the filtered test data sets and computed the proportion of mislabeled test observations to estimate the classification error rate for each classifier. Repeating this process 50 times, we computed the average of the error-rate estimates for each classifier. We remark that our simulation design avoids the selection bias discussed in \cite{Ambroise:2002fa}.

We used version 2.15.1 of the open-source statistical software {\tt R} for the classification study. For each classifier's {\tt R} implementation, we used the default settings with the exception that we explicitly set the prior probabilities as equal. In our {\tt R} implementation of the \emph{GRDA} classifier in the {\tt regdiscrim} package, we set $\alpha_k = 1$, $k = 1, \ldots, K$ for simplicity. We estimated $\lambda$ from a grid of 21 equidistant candidate values between 0 and 1, inclusively. Similarly, we selected $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using 10-fold cross-validation \citep{Hastie:2008dt}.

@@
